---
title: "**MATH 3190 Homework 6**"
subtitle: |
  Due March 19, 2026\
  Focus: Notes 7 Part 2\
  Total Points: 90 + 10 (Self-Critique Comment and Uploading to GitHub) = 100
format: 
  html:
    embed-resources: true
urlcolor: blue
execute:
  eval: true
  echo: true
editor_options: 
  chunk_output_type: console
---

Note that your homework should be completed in Quarto. You can just add your answers and code to this document in the appropriate part. The file should be rendered as an html document or pdf document (html is quicker to render and you will need an installation of LaTeX to render as a pdf). You will "turn in" this homework by uploading the files, the .qmd and output file, to your GitHub `Math_3190_Assignment` repository in the `Homework/Homework_6` directory. Make sure all supporting files (if there are any) are in this directory as well. You should use `git` and `Unix` commands in a terminal for this (i.e. don't manually upload files to GitHub). In addition, you will submit just your html or pdf file on Canvas to unlock the solutions.

Don't forget to also put a [self-critique comment]{.underline} on Canvas indicating what you did correctly and incorrectly and include any questions you may have about the assignment after submitting the assignment and thoroughly looking through the solutions.

# Problem 1 (28 points)

An automobile consulting company wants to understand the factors on which the pricing of cars depends. The dataset `car_price_prediction` in the `math3190package` (and also found in the GitHub data folder) has information on the sales of 4340 vehicles.

### Part a (3 points)

Create new variables in the dataset by taking the natural log of selling price, year, and km_driven. This will help linearize the association between variables. Then fit a linear model for predicting the log of selling price using all other variables except "name", "year", "selling_price", and "km_driven" (the logged variables should be used instread of those last three). Also, there is only one vehicle with a fuel type of "Electric". Remove that vehicle from the dataset.

```{r}
library(math3190package)
library(dplyr)

data(car_price_prediction)

car_data <- car_price_prediction %>%
  filter(fuel != "Electric")

car_data <- car_data %>%
  mutate(log_selling_price = log(selling_price),
         log_year = log(year),
         log_km_driven = log(km_driven))

model <- lm(log_selling_price ~ . - name - selling_price - year - km_driven, data = car_data)

summary(model)
```

### Part b (5 points)

Now fit some LASSO models for predicting log selling price using all but the "name" variable. Try the following values for the regularization parameter: $\lambda=0,\ 0.01,\ 0.1,$ and $1$ and comment on how the coefficients of the model change.

Note: when $\lambda=0$, the LASSO coefficients should equal the OLS model coefficients. However, they will actually be a bit off here. That is because the `glmnet()` function has a `thresh` argument that sets a threshold for convergence. It is, by default, set to `1e-07`. To make the parameters match, you can change that to `thresh = 1e-14` instead. This is not necessary, though.

```{r}
library(glmnet) |> suppressPackageStartupMessages()
library(glmnetUtils) |> suppressPackageStartupMessages()
library(faraway) |> suppressPackageStartupMessages()

lasso_data <- car_data |> select(-name)

lasso_0 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1, lambda = 0)
lasso_001 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1, lambda = 0.01)
lasso_01 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1, lambda = 0.1)
lasso_1 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1, lambda = 1)

coef(lasso_0)
coef(lasso_001)
coef(lasso_01)
coef(lasso_1)

lasso_full <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1)
plot(lasso_full, xvar = "lambda")
```

As lambda increases, the coefficients get smaller. When lambda is 0, there is no penalty, so the model behaves similarly to OLS. When lambda is 0.01 and 0.1, the coefficients start to shrink, and the smaller ones get close to zero. When lambda is 1, all of the coefficients are zero.

### Part c (4 points)

Now fit some ridge regression models for predicting log selling price using all but name. Try the following values for the regularization parameter: $\lambda=0,\ 0.01,\ 0.1,$ and $1$ and comment on how the coefficients of the model change.

```{r}
ridge_0 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 0, lambda = 0)
ridge_001 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 0, lambda = 0.01)
ridge_01 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 0, lambda = 0.1)
ridge_1 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 0, lambda = 1)

coef(ridge_0)
coef(ridge_001)
coef(ridge_01)
coef(ridge_1)
```

Coefficients shrink toward zero as lambda increases, but it doesn't set the coefficients exactly equal to zero as the LASSO did. As lambda increases from 0 to 0.01 and 0.1, the coefficients shrink a bit, but all the predictors stay in the model. When lambda is 1, the coefficients are closer to zero, but none are exactly zero.

### Part d (4 points)

Now fit some elastic net regression models for predicting log price using all but name. Try the following values for the regularization parameter: $\lambda=0,\ 0.01,\ 0.1,$ and $1$ for $\alpha=1/3$ and then comment on how the coefficients of the model change for each $\alpha$ from parts (b), (c), and (d).

```{r}
enet_0 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1/3, lambda = 0)
enet_001 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1/3, lambda = 0.01)
enet_01 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1/3, lambda = 0.1)
enet_1 <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1/3, lambda = 1)

coef(enet_0)
coef(enet_001)
coef(enet_01)
coef(enet_1)
```

When lambda is 0, the coefficients are still similar to the OLS estimates. As lambda increases to 0.01 and 0.1, the coefficients shrink towards zero. Compared to ridge regression, some coefficients shrink more quickly. Compared to LASSO, there are fewer coefficients that are exactly zero. When lambda is 1, some coefficients become zero, but not as many as in the LASSO.

### Part e (4 points)

Use the `cv.glmnet()` to find the "optimal" $\lambda$ value (let's use the "lambda.1se") for LASSO ($\alpha=1$), ridge ($\alpha=0$), and elastic net (let's again use $\alpha = 1/3$) and then fit models using the `glmnet()` function for all three using their respective "best" $\lambda$. **Set a seed** before running the `cv.glmnet()` each time.

Compare each model's variables and coefficients.

```{r}
set.seed(3190)
cv_lasso <- cv.glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1)
best_lambda_lasso <- cv_lasso$lambda.1se
fit_lasso_best <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1, lambda = best_lambda_lasso)
coef(fit_lasso_best)

set.seed(2026)
cv_ridge <- cv.glmnet(log_selling_price ~ ., data = lasso_data, alpha = 0)
best_lambda_ridge <- cv_ridge$lambda.1se
fit_ridge_best <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 0, lambda = best_lambda_ridge)
coef(fit_ridge_best)

set.seed(6767)
cv_enet <- cv.glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1/3)
best_lambda_enet <- cv_enet$lambda.1se
fit_enet_best <- glmnet(log_selling_price ~ ., data = lasso_data, alpha = 1/3, lambda = best_lambda_enet)
coef(fit_enet_best)
```

For LASSO, some coefficients are exactly zero, which removes variables deemed less important. Ridge keeps all variables in the model, but shrinks the coefficients toward zero. Elastic net is in between the two, some coefficients are zero, but not as many as LASSO, and the remaining coefficients are shrunk.

### Part f (8 points)

Use bootstrapping with at least 1000 samples to estimate the standard errors of the coefficients of the ridge regression model. For each bootstrap sample, run the `cv.glmnet()` function to find the best $\lambda$ and fit a model using that optimized $\lambda$. Use the "lambda.1se" for this. This can take a few minutes to run. Then compare these bootstrapped standard errors to the standard errors for the OLS model you fit in part a. Are they larger or smaller? Does this make sense?

```{r}
set.seed(3190)
B <- 1000

ridge_coefs <- replicate(B, {
  boot_data <- lasso_data[sample(1:nrow(lasso_data), replace = TRUE), ]
  
  cv_fit <- cv.glmnet(log_selling_price ~ ., data = boot_data, alpha = 0)
  best_lambda <- cv_fit$lambda.1se
  
  fit <- glmnet(log_selling_price ~ ., data = boot_data, alpha = 0, lambda = best_lambda)
  
  as.numeric(coef(fit))
})

ridge_coefs <- t(ridge_coefs)

ridge_se <- apply(ridge_coefs, 2, sd)
ridge_se

ols_coefs <- coef(model)
ols_se <- summary(model)$coefficients[, "Std. Error"]

names_ridge <- rownames(coef(fit_ridge_best))
ols_se_full <- rep(NA, length(names_ridge))
names(ols_se_full) <- names_ridge

common <- intersect(rownames(summary(model)$coefficients), names_ridge)
ols_se_full[common] <- summary(model)$coefficients[common, "Std. Error"]

data.frame(Coefficient = names_ridge, OLS_SE = ols_se_full, Ridge_Bootstrap_SE = ridge_se)
```

Compared to OLS, most of the ridge SEs are smaller. This makes sense because ridge shrinks coefficients toward zero, which reduces their variability.

# Problem 2 (21 points)

In 1990, sixty districts in California were randomly selected with the goal of predicting the median house value for the district based on the district location (longitude and latitude), the median house age in the block group, the total number of rooms in the homes, the total number of bedrooms, the population in the district, the number of households, and the median income (in \$10,000). The data are already in the .qmd file.

```{r}
#| label: house_data
#| echo: false
library(tidyverse) |> suppressPackageStartupMessages()
housing_train <- tibble(
  house_value = c(156100, 145300, 150000, 50600, 176300, 178500, 123800,
                  173900, 422700, 164400, 42100, 157800, 162900, 90500,
                  191800, 128100, 109700, 206700, 132200, 248200, 336900,
                  58000, 271500, 259500, 69400, 132700, 461200, 245000,
                  218000, 311700, 361600, 250000, 113900, 103600, 81900,
                  351900, 114100, 500001, 153300, 466700),
  longitude = c(-121.08, -121.27, -117.7, -119.29, -118.09, -117.67, 
                -120.76, -118.16, -118.39, -123.1, -117.66, -118.16, 
                -120.98, -120.99, -122.69, -118.18, -122.27, -122.11, 
                -121.99, -118.02, -118.02, -120.94, -118.6, -121.95, 
                -120.62, -118.18, -122.13, -117.93, -121.85, -118.49, 
                -122.44, -121.32, -117.07, -121.83, -121.43, -122.11, 
                -118.27, -118.4, -120.28, -117.16),
  latitude = c(38.95, 38.7, 34.06, 36.34, 33.9, 34.02, 38.47, 34.02,
               33.89, 38.79, 35.63, 33.97, 38.99, 37.67, 38.35, 33.91,
               37.83, 37.99, 38.34, 33.74, 33.73, 40.14, 34.26, 37.96, 
               36.99, 33.8, 37.4, 33.94, 36.6, 34.18, 37.76, 37.67, 32.74, 
               38, 38.53, 37.41, 34.01, 34.41, 37.9, 32.74),
  age = c(18, 16, 25, 28, 33, 16, 17, 44, 38, 20, 33, 23, 17, 28, 16, 41, 
          51, 16, 16, 26, 24, 31, 18, 18, 32, 42, 29, 30, 21, 31, 52, 21, 
          38, 15, 36, 27, 47, 22, 17, 43),
  rooms = c(1931, 3747, 2054, 1440, 3326, 3042, 1521, 1218, 1851, 3109, 
            2579, 1516, 3403, 1768, 1689, 1260, 2665, 3913, 1470, 3842, 
            6393, 3127, 6154, 2739, 2455, 2301, 6027, 2658, 2381, 3073, 
            2959, 1494, 1901, 6365, 2430, 5110, 921, 4443, 1047, 1437),
  bedrooms = c(380, 586, 609, 431, 720, 524, 309, 374, 332, 712, 564, 
               457, 661, 423, 254, 299, 574, 710, 261, 609, 1141, 664, 
               1070, 393, 508, 621, 1195, 382, 701, 674, 683, 271, 392, 
               1646, 426, 1599, 264, 560, 212, 406),
  population = c(1271, 1817, 2271, 2178, 2533, 1516, 607, 1175, 750, 1643,
                 1155, 1977, 1540, 1066, 921, 1535, 1258, 1782, 748, 1961,
                 2743, 1345, 3010, 1072, 1344, 2114, 2687, 1135, 1264, 1486,
                 1145, 781, 1099, 3838, 1199, 2764, 881, 1573, 530, 692),
  households = c(377, 590, 564, 440, 689, 475, 240, 342, 314, 638, 431, 435,
                 622, 392, 270, 322, 536, 676, 256, 595, 1057, 580, 1034, 374,
                 492, 561, 1171, 392, 659, 684, 666, 255, 406, 1458, 437,
                 1482, 221, 496, 196, 379)
)

housing_test <- tibble(
  house_value = c(214900, 44000, 194400, 128700, 435000, 179400, 129300,
                  312300, 76300, 225800, 295500, 162500, 340800, 93800,
                  155000, 283200, 133400, 248200, 265000, 57500),
  longitude = c(-119.25, -121.93, -122.31, -121.27, -118.42, -118.71, 
                -117.31, -122.04, -119.69, -117.88, -117.25, -121.48, 
                -122.49, -121.49, -116.18, -119.19, -118.23, -118.35, 
                -118.35, -124.15),
  latitude = c(34.26, 41.86, 38.27, 38.67, 33.85, 34.27, 34.02, 37.04, 
               36.79, 33.7, 32.8, 38.55, 37.74, 39.52, 33.69, 34.3, 33.89,
               34.28, 33.91, 40.88),
  age = c(30, 28, 34, 15, 43, 26, 18, 17, 15, 18, 26, 52, 48, 25, 17, 
          25, 16, 30, 31, 33),
  rooms = c(2948, 4225, 1748, 1701, 1584, 990, 1634, 4977, 2524, 2135, 
            2442, 2037, 1186, 848, 89, 2197, 5003, 3214, 2583, 2235),
  bedrooms = c(827, 835, 284, 346, 477, 223, 274, 994, 451, 373, 659, 
               358, 213, 153, 19, 320, 1180, 513, 663, 506),
  population = c(1635, 1908, 783, 723, 799, 719, 899, 1987, 1207, 1464, 
                 1134, 811, 487, 436, 79, 934, 4145, 1700, 1675, 1165),
  households = c(750, 686, 303, 352, 433, 232, 285, 947, 424, 405, 624, 
                 375, 207, 155, 21, 330, 1159, 533, 612, 441)
)
```

### Part a (2 points)

Using the training set, fit a linear regression model predicting house value from the other variables.

```{r}
house_lm <- lm(house_value ~ longitude + latitude + age + rooms + bedrooms + population + households, data = housing_train)

summary(house_lm)
```

### Part b (4 points)

Fit a ridge regression using `cv.glmnet()` to choose the optimal $\lambda$ using "lambda.min". Set the seed before using`cv.glmnet()`.

```{r}
set.seed(3190)

cv_ridge <- cv.glmnet(house_value ~ ., data = housing_train, alpha = 0)

best_lambda_ridge <- cv_ridge$lambda.min
best_lambda_ridge

ridge_model <- glmnet(house_value ~ ., data = housing_train, alpha = 0, lambda = best_lambda_ridge)

coef(ridge_model)
```

### Part c (5 points)

Compute the sum of squared errors (that is $\sum_{i=1}^n (y_i-\hat{y}_i)^2$ for both the OLS model and the ridge regression model for the **training set**. The `predict()` functions will be useful here. Compare the results and comment on why the OLS performs better here (the OLS should indeed perform better).

```{r}
y_train <- housing_train$house_value

pred_ols_train <- predict(house_lm, newdata = housing_train)
sse_ols_train <- sum((y_train - pred_ols_train)^2)
sse_ols_train

pred_ridge_train <- predict(ridge_model, newdata = housing_train)
sse_ridge_train <- sum((y_train - pred_ridge_train)^2)
sse_ridge_train
```

The results show OLS performs better. OLS fits the training data as close as possible with least squares, so it usually has smaller SSE on the training set. Ridge shrinks coefficients, so SSE is slightly higher.

### Part d (3 points)

Compute the sum of squared errors (that is $\sum_{i=1}^n (y_i-\hat{y}_i)^2$ for both the OLS model and the ridge regression model for the **testing set**. Remember to use the exact models you fit to the training sets. The `predict()` functions will be useful here. Compare the results and comment on why the ridge regression performs better here.

```{r}
y_test <- housing_test$house_value

pred_ols_test <- predict(house_lm, newdata = housing_test)
sse_ols_test <- sum((y_test - pred_ols_test)^2)
sse_ols_test

pred_ridge_test <- predict(ridge_model, newdata = housing_test)
sse_ridge_test <- sum((y_test - pred_ridge_test)^2)
sse_ridge_test
```

The results show the ridge regression does perform better. On the testing set, ridge usually performs better than OLS because it reduces variance and prevents overfitting. OLS might fit training data well but can overfit it, which leads to higher SSE on new data.

### Part e (7 points)

Use the `train()` function in the `caret()` package to find the optimal $\alpha, \lambda$ pair in this situation (using the training set and **10-fold cross validation**). For $\lambda$, search for values between 0 and 10000 by 100 and for $\alpha$, search for values between 0 and 1 by 0.05. Then fit a model using the optimal $\alpha$ and $\lambda$ using `glmnet()` and compare the sum of squared errors using this optimized model on the test set to the two models in part (d). Set a seed before running the `train()` function.

```{r}
library(caret) |> suppressPackageStartupMessages()
library(glmnet)

set.seed(3190)

tune_grid <- expand.grid(alpha = seq(0, 1, by = 0.05), lambda = seq(0, 10000, by = 100))

enet_model <- train(house_value ~ ., data = housing_train, method = "glmnet", tuneGrid = tune_grid, trControl = trainControl(method = "cv", number = 10))

enet_model$bestTune

best_alpha <- enet_model$bestTune$alpha
best_lambda <- enet_model$bestTune$lambda

final_enet <- glmnet(house_value ~ ., data = housing_train, alpha = best_alpha, lambda = best_lambda)

pred_enet_test <- predict(final_enet, newdata = housing_test)
sse_enet_test <- sum((y_test - pred_enet_test)^2)
sse_enet_test
```

The elastic net model finds the best mix between ridge and LASSO. This SSE in between the two SSE from part (d). It is a little bigger than the ridge, and less than the OLS.

# Problem 3 (41 points)

In the `AER` package is the data set "ShipAccidents". You can install that package and load that data using `data(ShipAccidents)`. Type `?ShipAccidents` to read about that data set.

### Part a (4 points)

Load in the data, remove the rows for which service is equal to 0, and then fit a Poisson regression model for predicting incidents from all other variables using the ShipAccidents data. Don't forget to put `family = "poisson"` in the `glm()` function. Briefly explain why it makes sense to do Poisson regression here.

```{r}
library(AER) |> suppressPackageStartupMessages()

data("ShipAccidents")

pois_model <- glm(incidents ~ ., data = ShipAccidents, family = "poisson")
pois_model
```

It makes sense to use Poisson regression here because the incidents is a count, and this type of model predicts counts that can't be negative. You can't have negative ship accidents.

### Part b (5 points)

Using that model, interpret the slope of "service" and the slope of "construction1970-74" in original (not log) units.

Ships constructed between 1970 and 1974 are expected to have around 4.94 times as many incidents as the reference level, holding all other variables constant.

### Part c (6 points)

Make a residual plot of the Pearson residuals vs the linear predictors and make a QQ plot of the Pearson residuals. Comment on what these imply about the fit.

```{r}
pearson_resid <- residuals(pois_model, type = "pearson")
plot(predict(pois_model), pearson_resid, xlab = "Linear predictor", ylab = "Pearson residuals")

qqnorm(pearson_resid)
```

The residual plot implies the model doesn't fit the data perfectly and there might be overdispersion.

### Part d (5 points)

Conduct a deviance goodness-of-fit (GOF) test for a lack of fit here. Type out the null and alternative hypotheses, report the test statistic, give the p-value, and provide an interpretation.

```{r}
deviance_stat <- deviance(pois_model)
df_resid <- df.residual(pois_model)
p_value <- 1 - pchisq(deviance_stat, df_resid)
deviance_stat
p_value
```

Null hypothesis: The Poisson model fits the data. Alternative hypothesis: The Poisson model does not fit the data. The test statistic is 77.44977, and the p-value is 1.542447e-07. Because the p-value is so small, we reject the null hypothesis, and the model does not fit well.

### Part e (4 points)

Is there evidence that this model has overdispersion? Explain.

```{r}
deviance(pois_model) / df.residual(pois_model)
```

The ratio is above one, so there is overdispersion. Overdispersion means the actual variance is greater than the mean. The residual plot also suggested overdispersion.

### Part f (6 points)

Fit a "quasipoisson" model to these data. Using this model, obtain an approximate 95% confidence interval for the mean response for a ship of type "B" with construction between 1965 and 1969, with operation between 1960 and 1974, and with 2000 aggregate months of service. You can construct this interval using a t critical value with 24 degrees of freedom since the quasipoisson family more closely follows a t distribution than a normal one. Interpret the interval.

```{r}
shipmodquasi = glm(incidents ~ ., data = ShipAccidents, family = "quasipoisson")
summary(shipmodquasi)
shippred = predict(shipmodquasi, data.frame(type = "B", construction = "1965-69", operation = "1960-74", service = 2000), se.fit = T)
exp(shippred$fit - qt(.975, 24)*shippred$fit + qt(.975, 24)*shippred$se.fit)
```

There is 95% confidence that the mean number of ship accidents for a ship with these characteristics lies between 0.31502 and 1.

### Part g (5 points)

Now construct an approximate 95% prediction interval for the number of incidents for the ship described in part f using the `qpois()` function. Interpret the interval.

```{r}
qpois(0.025, 0.31502)
qpois(0.975, 1)
```

### Part h (6 points)

Suppose we did not catch the fact that this model is overdispersed. Repeat parts f and g using the model fit with the "poisson" family, not "quasipoisson" and using a z interval for the mean response rather than a t interval. Compare your results to parts f and g and comment on what is different.

```{r}
shipmodpoisson = glm(incidents ~ ., data = ShipAccidents, family = "poisson")
summary(shipmodquasi)
shippred = predict(shipmodquasi, data.frame(type = "B", construction = "1965-69", operation = "1960-74", service = 2000), se.fit = T)
exp(shippred$fit - 1.96*shippred$fit + 1.96*shippred$se.fit)
```

The upper bound was still 1 for this, and the lower bound was just slightly higher at 0.3671551. This model gives a interval.
