---
title: "**MATH 3190 Homework 7**"
subtitle: |
  Due March 26, 2026\
  Focus: Notes 8\
  Total Points: 90 + 10 (Self-Critique Comment and Uploading to GitHub) = 100
format: 
  html:
    embed-resources: true
urlcolor: blue
execute:
  eval: true
  echo: true
editor_options: 
  chunk_output_type: console
---

Note that your homework should be completed in Quarto. You can just add your answers and code to this document in the appropriate part. The file should be rendered as an html document or pdf document (html is quicker to render and you will need an installation of LaTeX to render as a pdf). You will "turn in" this homework by uploading the files, the .qmd and output file, to your GitHub `Math_3190_Assignment` repository in the `Homework/Homework_7` directory. Make sure all supporting files (if there are any) are in this directory as well. You should use `git` and `Unix` commands in a terminal for this (i.e. don't manually upload files to GitHub). In addition, you will submit just your html or pdf file on Canvas to unlock the solutions.

Don't forget to also put a [self-critique comment]{.underline} on Canvas indicating what you did correctly and incorrectly and include any questions you may have about the assignment after submitting the assignment and thoroughly looking through the solutions.

# Problem 1 (57 points)

In Homework 6, we briefly looked at a small sample from this data set. Now we have the full data set with 20,640 districts. For districts in California from 1990, we want to predict the median house value for the district based on the district location (longitude and latitude), the median house age in the block group, the total number of rooms in the homes, the total number of bedrooms, the population in the district, the number of households, and the median income (in \$10,000). These data were obtained from [scikit-learn.org](https://scikit-learn.org/stable/datasets.html) and can be found in the Data folder of the Math3190_Sp26 GitHub repo with the name `cal_housing.data`.

### Part a (5 points)

Read in the data. The `read.csv()` function works for files that end in `.data` if they are comma separated like this file is. However, be aware that the housing data does not have column names, so make sure to put `header = FALSE`. Assign appropriate column names to the columns and then take logs (base e) of all variables except lat and long. Once this is done, split the data into training and testing sets with the testing containing 20% of the values. Set a seed when you do this.

```{r}
library(dplyr) |> suppressPackageStartupMessages()
library(caret) |> suppressPackageStartupMessages()
set.seed(3190)

housing_data <- read.csv("/Users/janewillden/Downloads/cal_housing.data", header = FALSE)

colnames(housing_data) <- c("longitude", "latitude", "median_house_age", "total_rooms", "total_bedrooms", "population", "households", "median_income", "median_house_value")

housing_log <- housing_data |>
  mutate(logage = log(median_house_age),
         logrooms = log(total_rooms),
         logbedrooms = log(total_bedrooms),
         logpop = log(population),
         loghouseholds = log(households),
         logincome = log(median_income),
         loghousevalue = log(median_house_value))

train_index <- createDataPartition(housing_log$loghousevalue, p = 0.8, list = FALSE)
housing_train <- housing_log[train_index, ]
housing_test <- housing_log[-train_index, ]
```

### Part b (5 points)

Use the following command to view pair-wise scatter plots for the data here. Change `housing_train` to whatever you called the training set and be sure to change the `eval` option for the code chunk to `true`. Note, that `lower` option changes the points to be plotted with periods instead of circles. This speeds up plotting time considerably and makes the plots more readable since there are many data points here.

Based on these plots, does multicollinearity appear to be an issue here? Explain.

```{r}
#| label: house_pairs
#| cache: true
#| eval: true
library(GGally)
ggpairs(housing_train, 
        lower = list(continuous = wrap("points", shape = "."))
)
```

Multicollinearity does seem to be an issue here. There is multiple pairs of predictors that seem to form linear patterns, which represents strong correlations between the variables. The plots where the points cluster closely along a line suggest that one variable can be predicted fairly easily from the other, which is an indication of multicollinearity.

### Part c (4 points)

Fit a OLS regression model predicting house prices from all other variables. Check its summary output and its VIF values. Comment on the VIFs.

```{r}
library(car) |> suppressPackageStartupMessages()

ols_fit <- lm(loghousevalue ~ logage + logrooms + logbedrooms + logpop + loghouseholds + logincome, data = housing_train)

summary(ols_fit)

vif_tbl <- tibble(Predictor = names(vif(ols_fit)), VIF = round(vif(ols_fit), 2))
print(vif_tbl)
```

The VIF shows that three predictors are highly collinear, and these are logrooms, logbedrooms, and loghouseholds. They all are greater than 10. On the other hand, logage and logincome have low VIFs which means they are mostly independent of other variables.

### Part d (3 points)

Fit a principal component model using the `pcr()` function in the `pls` package. Since the variables are on very different scales, use the `scale = TRUE` option in the function. Then find the VIF values for this model. The `vif()` function from the `car` library won't work here. Instead, you can take the diagonals of the inverse of the correlation matrix (like is shown in the regularization section of Notes 7) for the `scores` output. The code for this is `diag(solve(cor(pca_model$scores)))`. Of course, you need to change `pca_model` to what you called your model.

```{r}
library(pls) |> suppressPackageStartupMessages()

housing_pcr <- pcr(loghousevalue ~ logage + logrooms + logbedrooms + logpop + loghouseholds + logincome, data = housing_train, scale = TRUE)

scores_mat <- housing_pcr$scores[, 1:6]
vif_pcr <- diag(solve(cor(scores_mat)))

tibble(PC = paste0("PC", 1:6), VIF = round(vif_pcr, 3))
```

### Part e (5 points)

Now take the summary of your principal component model. With PCA, the common amount of variation we want to explain in the predictors is 90% or more. How many components are needed to achieve this? For that number of components, how much of the variation in log of home values is explained? How many components are needed to explain a "good" amount of the variation in the log of home values? Note: the upper bound on the amount of variation in the log of home values here with all 8 components will be equal to the $R^2$ value of the OLS model.

```{r}
explained_prop <- explvar(housing_pcr) / 100
total_explained <- cumsum(explained_prop)

tibble(PC = seq_along(total_explained), TotalVar = round(total_explained, 3)) |> print()
```

Three components are needed to explain at least 90% of the variance -- after PC 3, 96.9% of the variance is explained. A large number of componenets are needed to explain a good amount of variation in the log of home values.

### Part f (4 points)

Now fit a partial least squares model using the `plsr()` function in the `pls` package with `scale = TRUE`. Then find the VIF values for this model like you did in part (d) and comment on them.

```{r}
housing_pls <- plsr(loghousevalue ~ logage + logrooms + logbedrooms + logpop + loghouseholds + logincome, data = housing_train, scale = TRUE, validation = "CV", segments = 10)

opt_ncomp <- selectNcomp(housing_pls, method = "onesigma")
cat("Selected components:", opt_ncomp, "\n")

scores_mat <- housing_pls$scores[, 1:opt_ncomp, drop = FALSE]

vif_pls <- diag(solve(cor(scores_mat)))
tibble(PC = paste0("PC", 1:opt_ncomp),
       VIF = round(vif_pls, 3)) |> print()
```

The VIFs for the PLS compoenents are all one which means that each component is uncorrelated with the others. There is no multicollinearity among them, which is different from the OLS model.

### Part g (4 points)

Now take the summary of your partial least squares model. Compare this output to the summary of the PCA model. Explain what the differences are.

How many components are needed to balance a good amount of variation explained in $X$ and in $y$ for the PLS model?

```{r}
summary(housing_pls)
```

The PCA summary explains variation in the predictors, while the PLS summary explains both the predictors and the response. About 4 components are needed in the PLS model to balance explaining variation in both the predictors and the response while keeping prediction error low.

### Part h (8 points)

In part (e), you said how many components were needed for the PCA to explain a good amount of the $Y$ variable. Using that number of components, use the PCA and PLS models to find the root MSE when predicting the testing set. Note: the root MSE (or residual standard error, also abbreviated RMSE) is found by taking the square root of the sum of squared residuals divided by the residual degrees of freedom, which is $n-p-1$, where $p$ is the number of predictors in the model. Also, be aware that when you use the `predict()` function on your model, it will give you predictions for each number of components. You can access the predictions for using 3 components, for example, by typing `predict(pca_model, test_set)[,,3]`.

Then in part (g), you said how many components were needed for the PLS Using that number of components. Using this number of components, use the PCA and PLS models to find the root MSE predicting the testing set.

Compare the results of the predictions in both cases.

```{r}
pca_ncomp <- 3
pls_ncomp <- 4

pred_pca_all <- predict(housing_pcr, newdata = housing_test, ncomp = pca_ncomp)
pred_pca <- as.vector(pred_pca_all)

pred_pls_all <- predict(housing_pls, newdata = housing_test, ncomp = pls_ncomp)
pred_pls <- as.vector(pred_pls_all)
n_test <- nrow(housing_test)

rmse_custom <- function(y_obs, y_pred, p){rss   <- sum((y_obs - y_pred)^2)
  df <- n_test - p - 1
  sqrt(rss / df)}

rmse_pca <- rmse_custom(housing_test$loghousevalue, pred_pca, p = pca_ncomp)
rmse_pls <- rmse_custom(housing_test$loghousevalue, pred_pls, p = pls_ncomp)

tibble(
  Model = c("PCA (PCR)", "PLS"),
  Components = c(pca_ncomp, pls_ncomp),
  RMSE_log = c(round(rmse_pca, 4), round(rmse_pls, 4)),
  RMSE_original = c(round(exp(rmse_pca) - 1, 2), round(exp(rmse_pls) - 1, 2)) ) |> print()
```

The test-set RMSE for the 3 comp PCR model is 0.412, while the 4 comp PLS model has a lower RMSE of 0.378. This means the PLS model reduces the prediction error.

### Part i (12 points)

Using the number of components you said were needed for the PLS model in part (g), come up with reasonable surrogates for each of those components. Look at the projection output for this. When creating your surrogates, make sure to center and scale each variable. You can do this with the `scale()` function and putting your housing data frame or tibble in it.

Then fit a OLS model using those surrogates in the training set. Check the VIF of this model to make sure each one is below 5. If they are not, your surrogates should be changed. You can use the `vif()` function in the `car` package here.

```{r}
X_raw <- housing_train |> 
  select(logage, logrooms, logbedrooms, logpop, loghouseholds, logincome)

X_scaled <- scale(X_raw)

load_mat <- housing_pls$loading.weights[, 1:pls_ncomp]

surrogates_mat <- X_scaled %*% load_mat

surrogates_tbl <- as_tibble(surrogates_mat, .name_repair = ~ paste0("Surrogate", 1:pls_ncomp))

train_with_surrogates <- housing_train |> bind_cols(surrogates_tbl)

ols_surrogate <- lm(loghousevalue ~ .,data = train_with_surrogates |> select(loghousevalue, starts_with("Surrogate")))

vif_vals <- vif(ols_surrogate)
```

The surrogates were constructed as linear combinations of the scaled predictors using the PLS loading weights, which represent the PLS components. The VIF values for all surrogate variables are below 5, suggesting that multicollinearity isn't an issue in this model.

### Part j (7 points)

Now using that model you fit in the previous part with the surrogates, find the root MSE for predicting the testing set using that model, and compare it to what you got in part (h). Describe some pros and cons of the surrogate model.

Hint: to predict using this model, you will have to create a `newdata` data frame (or tibble) and then redefine your surrogates in that data frame using the scaled testing data set. Make sure the variable names in the `newdata` data frame match the variables names used in the model you defined in part (i).

```{r}
X_test_raw <- housing_test |> 
  select(logage, logrooms, logbedrooms, logpop, loghouseholds, logincome)

X_test_scaled <- scale(X_test_raw, center = attr(X_scaled, "scaled:center"), scale  = attr(X_scaled, "scaled:scale"))

surrogates_test <- X_test_scaled %*% load_mat

surrogates_test_tbl <- as_tibble(surrogates_test, .name_repair = ~ paste0("Surrogate", 1:pls_ncomp))

pred_surrogate <- predict(ols_surrogate, newdata = surrogates_test_tbl)

rmse_surrogate <- rmse_custom(housing_test$loghousevalue, pred_surrogate, p = pls_ncomp)
```

Pros: low VIF -- removes multicollinearity, can use standard OLS tools Cons: Less interpretable than original variables, requires extra preprocessing The RMSE of the surrogate model is very similar to the PLS model and lower than the PCR model, which suggests that it retains most of the predictive power of PLS.

# Problem 2 (33 points)

In base **R**, there is a dataset called `penguins`. Type `?penguins` to read a little about the data.

### Part a (6 points)

Remove the rows in the penguins dataset that have missing values. Then use the `ggpairs()` function from `GGally` to make pairwise scatterplots for the four quantitative variables and color the points based on the `species` variable. Note that the `ggpairs()` function has a `columns` argument where you can select the columns you want to plot and it has a `mapping` argument where you can put `aes(color = species, alpha = 0.5)` to color the points by species.

Include the plot here and then comment on which scatterplots seem to nicely separate the penguins by species and which tend to have worse separation.

```{r}
library(ggplot2)
library(GGally)
library(dplyr)

data(penguins)

penguins_clean <- penguins |> na.omit()

ggpairs(penguins_clean, columns = 3:6, mapping = aes(color = species, alpha = 0.5))
```

The scatterplots of flipper length vs. body mass and bill length vs. bill depthshow strong separation between speccies. In particular, Gentoo penguins tend to have a larger body mass and flipper length, which makes them more clearly distinguishable. The scatterplots with bill length and bill depth show strong separation between Adelie and Chisntrap species. There are some plots, (ex: jsut body mass or flipper length), show more overlap between Adelie and Chinstrap penguins, which suggests weaker separation in those cases.

### Part b (4 points)

Run a scaled principal component analysis on the four quantitative variables using the `prcomp()` function. How much of the total variation in these variables is explained by the first two principal components? You can use the `summary()` function on your PCA to find this easily.

```{r}

penguins_scaled <- scale(penguins_clean[, 3:6])

penguins_pca <- prcomp(penguins_scaled)

summary(penguins_pca)
```

The first two PCs explain about 88% of the total variation in the data.

### Part c (6 points)

Plot the first two principal components with PC1 on the x-axis and PC2 on the y-axis. Color the points by species and comment on how well the first two principal components are separating by species.

```{r}
pca_df <- as.data.frame(penguins_pca$x)
pca_df$species <- penguins_clean$species

ggplot(pca_df, aes(x = PC1, y = PC2, color = species)) +
  geom_point(alpha = 0.7) +
  labs(title = "PCA of Penguins Data")
```

The first two principal components give pretty good separation between species. Gentoo penguins are clearly separated from the other two species, mostly along PC1. Adelie and Chinstrap penguins show a bit overlap, though there'ss still partial separation along PC2. PCA captures most of the structure in the data, but doesn't perfectly separate all species.

### Part d (12 points)

Now use the `umap()` function in the `umap` library to fit a UMAP to the **scaled** quantitative variables in the penguins dataset. You can use the `scale()` function to scale the variables. Make sure the UMAP gives two components and set a `random_state` in the function so your results are reproducible. Then plot the UMAP using the the `umap_plot()` function in `math3190package` and color the points by `species` using the `color_umap` argument. Experiment with at least three different number of neighbors (`n_neighbors`) and at least three different minimum distances (`min_dist`).

```{r}
library(umap)
library(math3190package)

X_scaled <- scale(penguins_clean[, 3:6])

umap1 <- umap(X_scaled, n_neighbors = 5, min_dist = 0.1, random_state = 3190)
umap_plot(umap1, data = penguins_clean, color_umap = "species")

umap2 <- umap(X_scaled, n_neighbors = 10, min_dist = 0.5, random_state = 3190)
umap_plot(umap2, data = penguins_clean, color_umap = "species")

umap3 <- umap(X_scaled, n_neighbors = 20, min_dist = 0.9, random_state = 3190)
```

```{r}
umap_plot(umap3, data = penguins_clean, color_umap = "species")
```

### Part e (5 points)

Comment on the UMAP plots you made in part (d). Include a comparison to the PCA plot regarding how well these species are separated.

The UMAP plots usually show better separation between species compared to the PCA plot. Specifically, UMAP typicall forms more distinct clusters. Smaller values of n_neighbors and min_dist produce tighter and more separated clusters, while larger values produce more spread-out and overlapping clusters. Compared to PCA, UMAP can show a nonlinear relationship in the data, which can result in better separation between species. The quality of separation depends on the choice of parameters.

Feel free to repeat this problem using a variable other than species! Sometimes the UMAP outperforms PCA at separating classes, sometimes it does a little worse, and sometimes there are both quite bad.
